Reduced plasticity and variance in physiological rates for terrestrial ectotherm populations under climate change

### 2024-01-17 ms.qmd
Figures and updated MS
Absract and start of intro

Climate change is expected to result in warmer and more variable thermal environments globally. Greater thermal variability is expeced to result in strong selection pressures leading to genetic adaptation and/or the evolution of adaptive phenotypic plasticity. However, such responses depend on genetic and phenotypic variability, yet most work has focused on plastic changes in mean responses to climate warming ignoring how temperature changes variability. Here, using a large database of physiological rate data on marine, freshwater and terrestrial ectotherms we quantify the capacity for acclimation and the extent to which relative variance in physiological rates change in response to higher temperatures. Using new effect size measures and meta-analysis we show that physiological acclimation and changes in relative variance do not depend on past temperature variability and predictability experienced by a population or differ between life stages. Across habitats there was a general tendency for relative variation in physiological rates to decrease at higher temperatures. Freshwater ectotherms are capable of acclimating and such responses will likely be beneficial considering that these effects are larger than reductions in relative variance in physiological rates. In contrast, acclimation responses of marine organisms coincide with a nearly equal reduction in relative variance in physiolgical rates, suggesting acclimation may not be as beneficial as previously thought. Relative variance reductions are particularly pronounced for terrestrial ectotherms and this coincides with a general inability to acclimate highlighting the vulnerability of terrestrial ectotherms to climate change. Our results highlight the importance of considering changes in variance in physiological rates when predicting the capacity of populations to persist in and adapt to novel environments.

@@

Climate change is expected to result in warmer and more variable thermal environments for many organisms globally [@Suarez2023; @Easterling2000; @Ummenhofer2017]. Greater thermal variability in the past is expected to have resulted in strong selection pressures leading to genetic adaptation and/or the evolution of adaptive phenotypic plasticity -- both of which are considered important for population resilience to contemporary human-induced climate change [@Seebacher2015-wi; @Merila2014; @Chevin2015; @Chevin2017; @Chevin2010; @Nunney2016]. The extent to which adaptive plasticity and genetic adaptation will allow organisms to adapt to new environmental conditions is not yet clearly established [@Chevin2017]. Without such responses, high extinction rates are predicted if organisms cannot track suitable habitats [@Cahill2012; @Nunney2016]. Phenotypic plasticity is expected to be the 'first line of defence' against changing climates by buying time for genetic adaptation to take place [i.e., the 'plasticity first hypothesis' - @WestEberhard2003; see also @Lande2009]. Phenotypic plasticity is predicted to evolve when environmental variability is high but predictable and the costs of plasticity are low [@Chevin2017; @Scheiner2020; @Nunney2016; @Dewitt1998; @Reed2010]. Despite this theoretical expectation, empirical support is scant [but see @Leung2020], likely because many organisms can behaviorally adjust micro-habitat selection to offset thermal stress, the costs of plasticity are high [@Chevin2015; @Dewitt1998], and/or the prediction is likely only supported for specific plastic responses (i.e., active and developmental plasticity) or life-history stages.

@@

### PRISMA Flow Diagram

>>>>

Climate change is expected to result in warmer and more variable thermal environments globally. Greater thermal variability is expected to result in strong selection pressures leading to genetic adaptation and/or the evolution of adaptive phenotypic plasticity. Such responses depend on genetic and phenotypic variability, yet most work has focused on plastic changes in mean responses to climate warming ignoring how temperature changes variability. Here, using a large database of physiological rate data on marine, freshwater and terrestrial ectotherms we quantify the capacity for acclimation and the extent to which relative variance in physiological rates change in response to higher temperatures. Using new effect size measures and meta-analysis we show that physiological acclimation and changes in relative variance do not depend on past temperature variability and predictability experienced by a population, or differ between life stages. Across habitats there was a general tendency for relative variation in physiological rates to decrease at higher temperatures. Freshwater ectotherms are capable of acclimating and such responses will likely be beneficial considering that these effects are larger than reductions in relative variance in physiological rates. In contrast, acclimation responses of marine organisms coincide with a nearly equal reduction in relative variance in physiolgical rates, suggesting acclimation may not be as beneficial as previously thought. Relative variance reductions are particularly pronounced for terrestrial ectotherms and this coincides with a general inability to acclimate highlighting the vulnerability of terrestrial ectotherms to climate change. Our results highlight the importance of considering changes in variance in physiological rates when predicting the capacity of populations to persist in and adapt to novel environments.

@@

Climate change is expected to result in warmer but also more variable thermal environments for many organisms globally [@Suarez2023; @Easterling2000; @Ummenhofer2017]. Greater thermal variability in the past is expected to have resulted in strong selection pressures leading to genetic adaptation and/or the evolution of adaptive phenotypic plasticity -- both of which are considered important for population resilience to contemporary human-induced climate change [@Seebacher2015-wi; @Merila2014; @Chevin2015; @Chevin2017; @Chevin2010; @Nunney2016]. The extent to which adaptive plasticity and genetic adaptation will allow organisms to adapt to new environmental conditions is not yet clearly established [@Chevin2017]. Without such responses, high extinction rates are predicted if organisms cannot track suitable habitats [@Cahill2012; @Nunney2016]. Phenotypic plasticity is expected to be the 'first line of defence' against changing climates by buying time for genetic adaptation to take place [i.e., the 'plasticity first hypothesis' - @WestEberhard2003; see also @Lande2009]. Phenotypic plasticity is predicted to evolve when environmental variability is high but predictable and the costs of plasticity are low [@Chevin2017; @Scheiner2020; @Nunney2016; @Dewitt1998; @Reed2010]. Despite this theoretical expectation, empirical support is scant [but see @Leung2020], likely because many organisms can behaviorally adjust micro-habitat selection to offset thermal stress, the costs of plasticity are high [@Chevin2015; @Dewitt1998], and/or the prediction is likely only supported for specific plastic responses (i.e., active and developmental plasticity) or life-history stages.

@@

#### PRISMA Flow Diagram
